---
title: "IPUMS Read"
output: html_document
---

# Aggregation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading in the raw data

```{r}
if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("usa_00002.xml")
data <- read_ipums_micro(ddi)
```

## Converting to tibble

```{r}
library(tibble)
library(dplyr)
library(tidyverse)
d = tibble(data)
```

## Consolidating Race
```{r}
d = d %>% select(-RACED, -HISPAND) 

# change race column such that "Hispanic" is a race (like it is treated in all American data analysis) 
# note we temporarily label as race `10` (cause apparently they divided Asian into like 3 races)
d = d %>% mutate(RACE = replace(RACE, HISPAN != 0, 10)) %>% select(-HISPAN)

# combining asians into race `4`
d = d %>% mutate(RACE = replace(RACE, RACE %in% c(5, 6), 4))

# combining Others/Mixed into race `7`
d = d %>% mutate(RACE = replace(RACE, RACE %in% c(8, 9), 7))
```

## Removing Unemployed/No income
```{r}
d = d %>% filter(INDNAICS != '0') %>% filter(INCWAGE != 0) %>% filter(INCWAGE != 999999) %>% filter(INCWAGE != 999998)
```

## Consolidating Industries
```{r}
naics = function(indnaics) {
  mono = substr(indnaics, 1, 1)
  bi = substr(indnaics, 1, 2)
  internal = function(mono, bi){
    case_when(
      mono == '1' ~ 'Agriculture',
      bi == '21' ~ 'Mining',
      bi == '22' ~ 'Utilities',
      bi == '23' ~ 'Construction',
      mono == '3' ~ 'Manufacturing',
      bi == '42' ~ 'Wholesale Trade',
      bi == '44' ~ 'Retail',
      bi == '45' ~ 'Retail',
      mono == '4' ~ 'Transportation',
      bi == '51' ~ 'Media',
      bi == '52' ~ 'Banking/Insurance',
      bi == '53' ~ 'Real Estate/Lending',
      bi == '54' ~ 'Science/Tech',
      bi == '55' ~ 'Management',
      bi == '56' ~ 'Administration',
      bi == '61' ~ 'Education',
      bi == '62' ~ 'Healthcare',
      bi == '71' ~ 'Arts',
      bi == '72' ~ 'Food/Lodging',
      mono == '8' ~ 'Others',
      mono == '9' ~ 'Government',
      #.default = 'None'
    )}
  return(internal(mono, bi))
}
d = d %>% mutate(INDUSTRY = naics(INDNAICS))
# removing INDAICS
d = d %>% select(-INDNAICS)
```

## Replacing w/ State Names
```{r}
d = d %>% mutate(STATE = case_when(
  STATEFIP == 1 ~ 'Alabama',
  STATEFIP == 2 ~ 'Alaska',
  
  STATEFIP == 4 ~ 'Arizona',
  STATEFIP == 5 ~ 'Arkansas',
  STATEFIP == 6 ~ 'California',
  
  STATEFIP == 8 ~ 'Colorado',
  STATEFIP == 9 ~ 'Connecticut',
  STATEFIP == 10 ~ 'Delaware',
  STATEFIP == 11 ~ 'District of Columbia',
  STATEFIP == 12 ~ 'Florida',
  STATEFIP == 13 ~ 'Georgia',
  
  STATEFIP == 15 ~ 'Hawaii',
  STATEFIP == 16 ~ 'Idaho',
  STATEFIP == 17 ~ 'Illinois',
  STATEFIP == 18 ~ 'Indiana',
  STATEFIP == 19 ~ 'Iowa',
  STATEFIP == 20 ~ 'Kansas',
  STATEFIP == 21 ~ 'Kentucky',
  STATEFIP == 22 ~ 'Louisiana',
  STATEFIP == 23 ~ 'Maine',
  STATEFIP == 24 ~ 'Maryland',
  STATEFIP == 25 ~ 'Massachusetts',
  STATEFIP == 26 ~ 'Michigan',
  STATEFIP == 27 ~ 'Minnesota',
  STATEFIP == 28 ~ 'Mississippi',
  STATEFIP == 29 ~ 'Missouri',
  STATEFIP == 30 ~ 'Montana',
  STATEFIP == 31 ~ 'Nebraska',
  STATEFIP == 32 ~ 'Nevada',
  STATEFIP == 33 ~ 'New Hampshire',
  STATEFIP == 34 ~ 'New Jersey',
  STATEFIP == 35 ~ 'New Mexico',
  STATEFIP == 36 ~ 'New York',
  STATEFIP == 37 ~ 'North Carolina',
  STATEFIP == 38 ~ 'North Dakota',
  STATEFIP == 39 ~ 'Ohio',
  STATEFIP == 40 ~ 'Oklahoma',
  STATEFIP == 41 ~ 'Oregon',
  STATEFIP == 42 ~ 'Pennsylvania',
  
  STATEFIP == 44 ~ 'Rhode Island',
  STATEFIP == 45 ~ 'South Carolina',
  STATEFIP == 46 ~ 'South Dakota',
  STATEFIP == 47 ~ 'Tennessee',
  STATEFIP == 48 ~ 'Texas',
  STATEFIP == 49 ~ 'Utah',
  STATEFIP == 50 ~ 'Vermont',
  STATEFIP == 51 ~ 'Virginia',
  
  STATEFIP == 53 ~ 'Washington',
  STATEFIP == 54 ~ 'West Virginia',
  STATEFIP == 55 ~ 'Wisconsin',
  STATEFIP == 56 ~ 'Wyoming',
))
d = d %>% select(-STATEFIP)
```

## Consolidating counts (PERWT) of condensed form

```{r}
d = d %>% group_by(SEX, AGE, RACE, INCWAGE, INDUSTRY, STATE) %>% summarize(COUNT = sum(PERWT)) %>% ungroup()
```

## Consolidating into age groups
```{r}
d = d %>% mutate(AGE_GROUP = case_when(
  0 <= AGE & AGE < 25 ~ 0,
  25 <= AGE & AGE < 35 ~ 25,
  35 <= AGE & AGE < 45 ~ 35,
  45 <= AGE & AGE < 55 ~ 45,
  55 <= AGE ~ 55
)) %>% select(-AGE) %>% 
  group_by(SEX, AGE_GROUP, RACE, INCWAGE, INDUSTRY, STATE) %>%
  summarize(COUNT = sum(COUNT)) %>% 
  ungroup()
```

# Creating end tables

```{r}
pivot_sex = function(d) {
  d %>% pivot_wider(names_from = SEX, values_from = MED_INC) %>% rename(MED_INC_M = `1`, MED_INC_F = `2`)
}
```

## Wages by State and Sex
```{r}
dmap = d %>% select(-AGE_GROUP, -RACE, -INDUSTRY) %>% 
  group_by(STATE, SEX, INCWAGE) %>%
  summarize(COUNT = sum(COUNT)) %>%
  ungroup() %>% 
  group_by(STATE, SEX) %>%
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>%
  ungroup()

write.table(dmap %>% pivot_sex, file = "dmap.csv", sep=",", row.names = F)
```

## Wages by State and Sex and Race
```{r}
dmaprace = d %>% select(-AGE_GROUP, -INDUSTRY) %>%
  group_by(STATE, SEX, RACE, INCWAGE) %>%
  summarize(COUNT = sum(COUNT)) %>%
  ungroup() 

nationwide = dmaprace %>% 
  group_by(SEX, RACE) %>% 
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>% 
  ungroup() %>%
  mutate(STATE = 'United States')
  
dmaprace = dmaprace %>%
  group_by(STATE, SEX, RACE) %>%
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>%
  ungroup() %>%
  bind_rows(nationwide)

raceInt2Label = function (RACE) {
  case_when(
    RACE == 1 ~ 'White', 
    RACE == 2 ~ 'Black/African American', 
    RACE == 3 ~ 'American Indian/Alaska Native', 
    RACE == 4 ~ 'Asian', 
    RACE == 7 ~ 'Mixed/Other', 
    RACE == 10 ~ 'Hispanic'
  )
}

write.table(dmaprace %>% 
              pivot_sex %>% 
              mutate(RACE = raceInt2Label(RACE)), 
            file = "dmaprace.csv", sep=",", row.names = F)
```

## Wages by State and Sex and Industry
```{r}
dmapindustry = d %>% select(-AGE_GROUP, -RACE) %>%
  group_by(STATE, SEX, INDUSTRY, INCWAGE) %>%
  summarize(COUNT = sum(COUNT)) %>%
  ungroup()

nationwide = dmapindustry %>% 
  group_by(SEX, INDUSTRY) %>% 
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>% 
  ungroup() %>%
  mutate(STATE = 'United States')
  
dmapindustry = dmapindustry %>%
  group_by(STATE, SEX, INDUSTRY) %>%
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>%
  ungroup() %>%
  bind_rows(nationwide)

write.table(dmapindustry %>% pivot_sex, file = "dmapindustry.csv", sep=",", row.names = F)
```

## Wages by State and Sex and Age Group
```{r}
dmapages = d %>% select(-INDUSTRY, -RACE) %>%
  group_by(STATE, SEX, AGE_GROUP, INCWAGE) %>%
  summarize(COUNT = sum(COUNT)) %>%
  ungroup()

nationwide = dmapages %>% 
  group_by(SEX, AGE_GROUP) %>% 
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>% 
  ungroup() %>%
  mutate(STATE = 'United States')
  
dmapages = dmapages %>%
  group_by(STATE, SEX, AGE_GROUP) %>%
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>%
  ungroup() %>%
  bind_rows(nationwide)

ages2labels = function(d) {
  case_when(
    d == 0 ~ "Under 25",
    d == 25 ~ "25-34",
    d == 35 ~ "35-44",
    d == 45 ~ "45-54",
    d == 55 ~ "55+"
  )
}

write.table(dmapages %>% pivot_sex %>% mutate(AGE_GROUP = ages2labels(AGE_GROUP)), file = "dmapages.csv", sep=",", row.names = F)
```

## Wages by Race relative to White Men
```{r}
# find median income of all (employed/working) white men
MED_WHITE = ( d %>% filter(RACE == 1, SEX == 1) %>% summarize(MED_WHITE = median(rep(INCWAGE, COUNT))) )[[1,1]]

dwhitemen = d %>% select(-STATE, -INDUSTRY) %>%
  group_by(SEX, RACE, INCWAGE) %>%
  summarize(COUNT = sum(COUNT)) %>%
  ungroup() %>%
  group_by(SEX, RACE) %>%
  summarize(MED_INC = median(rep(INCWAGE, COUNT))) %>%
  ungroup() %>%
  mutate(RACE = raceInt2Label(RACE)) %>%
  arrange(MED_INC)

write.table(dwhitemen, file = "dwhitemen.csv", sep=",", row.names = F)
```
